Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-06-26 NovelLBM.tex
include cell render
update responses

To understand the rheology of such complex liquids in non-trivial geometries, computer simulations are an invaluable tool.
While a number of numerical techniques have been developed, the simulation of viscoelastic liquids remains an involved subject \cite{Alves.2020}, in particular for high viscosity ratios.
A very popular simulation technique for Newtonian fluids is the Lattice Boltzmann method (LBM) \cite{Kruger_book} due to its simplicity and high parallelizability.
LBM has also been extended to viscoelastic situations, mainly in two \cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005} but occasionally also in three \cite{malaspinas2010,gupta2015,gupta2015_2,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016} dimensions.
\RR{We will discuss these approaches in detail in section~\ref{sec:existingLBM} below.}
To alleviate stability problems, some of these works include an artificial diffusivity of the polymer stress \cite{malaspinas2010,ma2020,wang2020,wang2021}, or perform special decompositions of the stress tensor \cite{dzanic2022}.
Despite these improvements, however, none of these methods has been applied to very high viscosity ratio fluids which are often used as cell carrier fluids in lab experiments or technological applications.

@@

We introduce the ratio between the polymer and the solvent viscosity \RR{and express it as a ratio between the viscosity at zero shear rate, $ \eta _0$, and the viscosity at infinite shear rate $ \eta _ \infty $}:
\begin{align}
	R_\eta = \frac{\eta_\text{p}}{\eta_\text{s}} = \frac{\eta_0-\eta_\infty}{\eta_\infty}		\label{eq:R_eta}
\end{align}
which we and others found to be an important parameter to determine numerical stability of LB simulations, especially if no artificial diffusivity is used \cite{malaspinas2010,dzanic2022_2}.
This ratio can similarly be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979} using the zero-shear viscosity $\eta_0$ and the infinite-shear viscosity $\eta_\infty$.
For our alginate solution we obtain $R_\eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $R_\eta$.
The \SI{0.49}{\percent} methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

@@

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book}, and we therefore \RR{only mention the key choices here. Further details can be found in the appendix~\ref{app:LBM} or in literature.}
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collision operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{lehmann2021, Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Non-periodic boundaries are modeled using the half-way bounce back algorithm. To model fixed inflow/outflow, bounce back boundaries using a Dirichlet boundary condition as described by Krüger \textit{et al.} \cite{Kruger_book} are used.

@@

The spatio-temporal dynamics of the polymer stress $\ma{\tau}$ is given by equation~(\ref{eq:polymerStress}) in the form of an advection-reaction equation.
We solve this equation by a finite-volume scheme running in parallel (with identical time step) to the LBM solver.\RR{ For this, the simulation volume is subdivided in $1\times1\times1$ (in LU) cubes centered around each lattice node. The average value of $\ma{\tau}$ within these volumes is takes as the value at the lattice node.}
Advection is handled using the corner transport upwind scheme\RR{ (CTU)} \cite{kuron2021}.\RR{ CTU considers the finite volume cell to be virtually displaced by $\vec{u}\Delta t$, where $\vec{u}$ is the local velocity and $\Delta t$ is the size of a time step. The resulting overlap with the neighboring cells (as a fraction of the cell volume) dictates what percentage of the polymer stress currently located at the cell is advected to the respective neighbor. This scheme prevents checkerboard instabilities.} \SG{ add two phrases about CTU which somewhat non-standard }\RK{done}
\RR{ The remaining derivatives are calculated using finite differences.}
At the walls\RR{, which coincide with the surfaces of the finite volumes}, a zero-flux boundary condition for the stress is used.
The source terms are integrated using the Euler method.

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
Specifically, we implement the LBM collision as follows.
\begin{equation}
	f_i\left(\vec{x}+\vec{c}_i\Delta t, t+\Delta t\right) - f_i\left(\vec{x}, t\right) = -\frac{\Delta t}{\tau_\text{r}}\left[f_i\left(\vec{x}, t\right) - f_i^\text{eq}\left(\vec{x}, t\right) + B_i\right].
\end{equation}
Here, $f_i$ are the populations, $f_i^\text{eq}$ are the equilibrium populations, $\vec{c_i}$ are the lattice velocities, $\Delta t$ is the time step, $\tau_\text{r}$ is the relaxation time and $B_i$ are the stress contributions defined as follows.
\begin{equation}
	B_i = w_i\left(\frac{\vec{c}_i\vec{c}_i}{2c_\text{s}^4}-\frac{\uT}{2c_\text{s}^2}\right):\ma{\tau\RR{^\prime}}.
\end{equation}
Here, $w_i$ are the lattice weights, $c_\text{s}$ is the lattice speed of sound and $\ma{\tau\RR{^\prime}}$ is the \RR{stress tensor to be coupled into the LBM. In the case of viscoelastic fluids this typically is the} polymer stress tensor. \RR{For our method this tensor is modified as described in the following section. }The colon denotes the Frobenius inner product.
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

The key difficulty for simulating many realistic viscoelastic fluids is the imbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_\eta$ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^3$ (see figures~\ref{fig:fit_alginate}, \ref{tab:bioinks} and table~\ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polymer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} = 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	\label{eq:shuffleStress}
\end{align}
with the transferred ("shuffled") stress
\begin{align}
	\ma{\tau}_\text{shuffle} = 2\alpha_\text{s}\eta_\text{p}\ma{D}
\end{align}
containing an arbitrary parameter $\alpha_\text{s}$.
Finally, this results in
\begin{align}
	\ma{\sigma} = 2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $\alpha_\text{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity.
The polymer stress determined by equation~\ref{eq:polymerStress} is reduced by the amount $2\alpha_\text{s}\eta_\text{p}\ma{D}$ at every time step before \RR{$\ma{\tau^\prime} = \ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}$} is coupled into the LBM.

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$ here and check different values further below.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD_3D}). The polymer stress is then initialized using these values.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
\RR{An artificial increase of the Reynolds number to reduce computational time is possible in principle, but is not employed here to bring out clearly the characteristics of our method.}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
The time evolution of the center-line velocity is shown in figure~\ref{fig:2DTimeEvolution} of the appendix.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
Using the simulated velocities $\vec{u_i}$ and the semi-analytical results $\vec{u}$. This gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved L1 error defined as follows is shown in figure~\ref{fig:2DvelErr} in the appendix.
\begin{equation}
	e_\text{L1} = \frac{\vec{u_i} - \vec{u}\left(y_i\right)}{\vec{u}\left(y_i\right)}
\end{equation}
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.
Similarly, the exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/cellInShear_sigma_vM.pdf}
		\caption{The von Mises stress around a cell in viscoelastic shear flow.}
		\label{fig:cellInShear_vM}
	\end{figure}

@@

	\RR{

@@

	For the forces used to drive the simple periodic flows and the forces from the IBM used to simulate cells, we use the algorithm presented by Guo et al. \cite{guo2002}.  \SG{ what are ''simple forces''? }\RK{as opposed to stresses which are generalized forces. -> removed the "simple"}

@@

	\begin{equation}\label{eq:D3Q19c}
		\begin{split}
			&\frac{\Delta t}{\Delta x}\vec{c}_i = \\
			&\begin{Bmatrix}
				0 & 1 & -1 & 0 & 0 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 \\
				0 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & -1 & 1 & 0 & 0 & 1 & -1 \\
				0 & 0 & 0 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 & -1 & 1 & -1 & 1
			\end{Bmatrix}
		\end{split}
	\end{equation}
	Where each column is a vector.
	\begin{equation}\label{eq:D3Q19w}
		w_i = \begin{cases}
			\frac{1}{3},\quad i=0\\
			\frac{1}{18},\quad 1 \leq i \leq 6\\
			\frac{1}{36},\quad \text{otherwise}
		\end{cases}
	\end{equation}
	}

>>>>

To understand the rheology of such complex liquids in non-trivial geometries, computer simulations are an invaluable tool.
While a number of numerical techniques have been developed, the simulation of viscoelastic liquids remains an involved subject \cite{Alves.2020}, in particular for high viscosity ratios.
A very popular simulation technique for Newtonian fluids is the Lattice Boltzmann method (LBM) \cite{Kruger_book} due to its simplicity and high parallelizability.
LBM has also been extended to viscoelastic situations, mainly in two \cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005} but occasionally also in three \cite{malaspinas2010,gupta2015,gupta2015_2,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016} dimensions.
We will discuss these approaches in detail in section~\ref{sec:existingLBM} below.
To alleviate stability problems, some of these works include an artificial diffusivity of the polymer stress \cite{malaspinas2010,ma2020,wang2020,wang2021}, or perform special decompositions of the stress tensor \cite{dzanic2022}.
Despite these improvements, however, none of these methods has been applied to very high viscosity ratio fluids which are often used as cell carrier fluids in lab experiments or technological applications.

@@

We introduce the ratio between the polymer and the solvent viscosity and express it as a ratio between the viscosity at zero shear rate, $\eta_0$, and the viscosity at infinite shear rate $\eta_\infty$:
\begin{align}
	R_\eta = \frac{\eta_\text{p}}{\eta_\text{s}} = \frac{\eta_0-\eta_\infty}{\eta_\infty}		\label{eq:R_eta}
\end{align}
which we and others found to be an important parameter to determine numerical stability of LB simulations, especially if no artificial diffusivity is used \cite{malaspinas2010,dzanic2022_2}.
This ratio can similarly be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979} using the zero-shear viscosity $\eta_0$ and the infinite-shear viscosity $\eta_\infty$.
For our alginate solution we obtain $R_\eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $R_\eta$.
The \SI{0.49}{\percent} methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

@@

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book}, and we therefore only mention the key choices here. Further details can be found in the appendix~\ref{app:LBM} or in literature.
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collision operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{lehmann2021, Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Non-periodic boundaries are modeled using the half-way bounce back algorithm. To model fixed inflow/outflow, bounce back boundaries using a Dirichlet boundary condition as described by Krüger \textit{et al.} \cite{Kruger_book} are used.

@@

The spatio-temporal dynamics of the polymer stress $\ma{\tau}$ is given by equation~(\ref{eq:polymerStress}) in the form of an advection-reaction equation.
We solve this equation by a finite-volume scheme running in parallel (with identical time step) to the LBM solver. For this, the simulation volume is subdivided in $1\times1\times1$ (in LU) cubes centered around each lattice node. The average value of $\ma{\tau}$ within these volumes is takes as the value at the lattice node.
Advection is handled using the corner transport upwind scheme (CTU) \cite{kuron2021}. CTU considers the finite volume cell to be virtually displaced by $\vec{u}\Delta t$, where $\vec{u}$ is the local velocity and $\Delta t$ is the size of a time step. The resulting overlap with the neighboring cells (as a fraction of the cell volume) dictates what percentage of the polymer stress currently located at the cell is advected to the respective neighbor. This scheme prevents checkerboard instabilities.
The remaining derivatives are calculated using finite differences.
At the walls, which coincide with the surfaces of the finite volumes, a zero-flux boundary condition for the stress is used.
The source terms are integrated using the Euler method.

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
Specifically, we implement the LBM collision as follows.
\begin{equation}
	f_i\left(\vec{x}+\vec{c}_i\Delta t, t+\Delta t\right) - f_i\left(\vec{x}, t\right) = -\frac{\Delta t}{\tau_\text{r}}\left[f_i\left(\vec{x}, t\right) - f_i^\text{eq}\left(\vec{x}, t\right) + B_i\right].
\end{equation}
Here, $f_i$ are the populations, $f_i^\text{eq}$ are the equilibrium populations, $\vec{c_i}$ are the lattice velocities, $\Delta t$ is the time step, $\tau_\text{r}$ is the relaxation time and $B_i$ are the stress contributions defined as follows.
\begin{equation}
	B_i = w_i\left(\frac{\vec{c}_i\vec{c}_i}{2c_\text{s}^4}-\frac{\uT}{2c_\text{s}^2}\right):\ma{\tau^\prime}.
\end{equation}
Here, $w_i$ are the lattice weights, $c_\text{s}$ is the lattice speed of sound and $\ma{\tau^\prime}$ is the stress tensor to be coupled into the LBM. In the case of viscoelastic fluids this typically is the polymer stress tensor. For our method this tensor is modified as described in the following section. The colon denotes the Frobenius inner product.
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

The key difficulty for simulating many realistic viscoelastic fluids is the imbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_\eta$ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^3$ (see figures~\ref{fig:fit_alginate}, \ref{tab:bioinks} and table~\ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polymer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} = 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	\label{eq:shuffleStress}
\end{align}
with the transferred ("shuffled") stress
\begin{align}
	\ma{\tau}_\text{shuffle} = 2\alpha_\text{s}\eta_\text{p}\ma{D}
\end{align}
containing an arbitrary parameter $\alpha_\text{s}$.
Finally, this results in
\begin{align}
	\ma{\sigma} = 2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $\alpha_\text{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity.
The polymer stress determined by equation~\ref{eq:polymerStress} is reduced by the amount $2\alpha_\text{s}\eta_\text{p}\ma{D}$ at every time step before $\ma{\tau^\prime} = \ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}$ is coupled into the LBM.

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$ here and check different values further below.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD_3D}). The polymer stress is then initialized using these values.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
An artificial increase of the Reynolds number to reduce computational time is possible in principle, but is not employed here to bring out clearly the characteristics of our method.
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
The time evolution of the center-line velocity is shown in figure~\ref{fig:2DTimeEvolution} of the appendix.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
Using the simulated velocities $\vec{u_i}$ and the semi-analytical results $\vec{u}$. This gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved L1 error defined as follows is shown in figure~\ref{fig:2DvelErr} in the appendix.
\begin{equation}
	e_\text{L1} = \frac{\vec{u_i} - \vec{u}\left(y_i\right)}{\vec{u}\left(y_i\right)}
\end{equation}
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.
Similarly, the exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.15]{assets/images/cellInShear.png}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/cellInShear_sigma_vM.pdf}
		\caption{Cell in viscoelastic shear flow. (a) The 3D shape of the deformed cell and (b) a 2D cross-section of the von Mises stress.}
		\label{fig:cellInShear_vM}
	\end{figure}

@@

@@

	For the forces used to drive the simple periodic flows and the forces from the IBM used to simulate cells, we use the algorithm presented by Guo et al. \cite{guo2002}.

@@

	\begin{equation}\label{eq:D3Q19c}
		\begin{split}
			&\frac{\Delta t}{\Delta x}\vec{c}_i = \\
			&\begin{Bmatrix}
				0 & 1 & -1 & 0 & 0 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 \\
				0 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & -1 & 1 & 0 & 0 & 1 & -1 \\
				0 & 0 & 0 & 0 & 0 & 1 & -1 & 0 & 0 & 1 & -1 & 1 & -1 & 0 & 0 & -1 & 1 & -1 & 1
			\end{Bmatrix}
		\end{split}
	\end{equation}
	Where each column is a vector.
	\begin{equation}\label{eq:D3Q19w}
		w_i = \begin{cases}
			\frac{1}{3},\quad i=0\\
			\frac{1}{18},\quad 1 \leq i \leq 6\\
			\frac{1}{36},\quad \text{otherwise}
		\end{cases}
	\end{equation}

### 2024-06-26 responses/Response1.tex
update responses

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{Submission ID \SG{ to fill }}
\rfoot{\thepage}
\setlength\parindent{0pt}

\newcommand{\SG}[1]{{\color{red}SG:~#1}}
\newcommand{\SJM}[1]{{\color{orange}SM:~#1}}
\newcommand{\SW}[1]{{\color{purple}SW:~#1}}

@@

We have added a new appendix E where we discuss the LB method in general as well as our specific implementation details. 
\SG{der Appendix ist an sich ok, aber sollte noch frueher und deutlicher auf unsere Implementierung eingehen, z.B. D3Q19 bereits dort erwähnen, wo du auf velocity sets eingehst. Beim Guo forcing konkret sagen, wofuer wir es verwenden (''simple forces'' ist hier zu vage). Bei der Dichte und der Viskositaet ist es nicht so wichtig, was man ''typically'' nimmt, sondern was du konkret in deinen Simulationen genommen hast. }

@@

Our choice of the finite-volume method is motivated primarily by its clarity and simplicity when implementing complex boundary conditions.
Furthermore, the finite-volume method matches well with the so-called corner-transport-upwind (CTU) scheme which we use for advection and for which earlier works \SG{ give ref of Kuron } reported increased stability.
To make these points clearer to the reader, we added many corresponding details to section 4.2 of the revised manuscript.
%Finite difference is a lot harder to implement (boundary conditions), has a lot more loose ends to tie (to assure conservation for example) and is less stable.

@@

\SG{ Die Antwort geht so nicht, es geht hier nicht um eine Grundsatzdikussion zu MRT sondern um eine konkrete Frage zu unserem System.
Es gibt drei Moeglichkeiten zur Antwort:

(i) Wir bedanken uns artig für den Vorschlag und sagen, dass wir es in Zukunft ausprobieren werden, aber dass es für diese Arbeit zu weit fuehrt.

(ii) Wir sagen, dass MRT hier nichts nuetzt. Dann brauchen wir aber konkrete Belege dafuer. Fuer welche Systeme wurde MRT ausprobiert? Was war der Effekt? Konnte die Stabilitaet fuer den klassischen Ansatz ohne viscosity shuffling durch MRT verbessert werden?

(iii) Wir argumentieren, dass andere in der Literatur MRT benutzt haben und dass es dort auch nichts gebracht hat. Deswegen machen wir uns nicht die Muehe, es auszuprobieren.

Man kann die Varianten auch kombinieren. Kannst du zu (ii) und (iii) mal raussuchen, was da der Stand ist?}

>>>>

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{Submission ID FLD-24-0080}
\rfoot{\thepage}
\setlength\parindent{0pt}

\newcommand{\SG}[1]{{\color{red}SG:~#1}}
\newcommand{\SJM}[1]{{\color{orange}SM:~#1}}
\newcommand{\SW}[1]{{\color{purple}SW:~#1}}
\newcommand{\RK}[1]{{\color{cyan}RK: #1}}

\usepackage[
backend=bibtex,
citestyle=numeric-comp,
maxbibnames=10,
sorting=none
]{biblatex}
\addbibresource{../assets/bib/bib.bib}

\usepackage[locale=US]{siunitx}
\sisetup{per-mode=fraction}
\sisetup{separate-uncertainty=true}

@@

We have added a new appendix E where we discuss the LB method in general as well as our specific implementation details.

@@

Our choice of the finite-volume method is motivated primarily by its clarity and simplicity when implementing complex boundary conditions.
Furthermore, the finite-volume method matches well with the so-called corner-transport-upwind (CTU) scheme which we use for advection and for which earlier works \cite{kuron2021} reported increased stability.
To make these points clearer to the reader, we added many corresponding details to section 4.2 of the revised manuscript.
%Finite difference is a lot harder to implement (boundary conditions), has a lot more loose ends to tie (to assure conservation for example) and is less stable.

@@

MRT is considered to improve numerical stability of time dependent or turbulent flow in high $Re$ cases \cite{dzanic2022, Kruger_book}, it is used for $Re = \num{100}$ \cite{chai2011} and above \cite{dhumieres2002}. As the problem discussed in the present work has very low $Re$, it would not be expected to improve stability here. Consequently, SRT is used by all authors we list in our literature overview except for Kuron et al. (TRT) \cite{kuron2021} and Gupta et al. (MRT) \cite{gupta2015}. Notably, while Gupta et al. use MRT their method achieves the lowest $max\left(R_\eta\right)$ in our overview. We do not want to imply, that MRT might be counterproductive here (Gupta et al. did not compare to SRT), but it is also clearly no remedy. That being said, given, that we have MRT implemented anyway, we of course tried it before coming up with the present algorithm and were not able to observe any improvement.

### 2024-06-26 responses/Response2.tex
update responses

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{Submission ID \SG{ to fill }}
\rfoot{\thepage}
\setlength\parindent{0pt}

@@

Aside from figures 3 and 5, which are planar geometries, all our systems are fully 3D.
This includes the cylindrical flow in figure 4. 
Even more important, it includes the RT-DC geometry shown in figure 7 and the fully three-dimensional cell in figure 8.
To bring out more clearly the latter aspect, we have added to figure 8 an illustration of our simulated system \SG{ denote current figure as 8b, and add a standard 3D Paraview picture as 8a. }\RK{Im Sinne von Bild von der deformierten Zelle? Oder meinst du ein anderes System wo man mehr sieht als einen unförmigen blob im nichts?}

>>>>

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{Submission ID FLD-24-0080}
\rfoot{\thepage}
\setlength\parindent{0pt}

@@

Aside from figures 3 and 5, which are planar geometries, all our systems are fully 3D.
This includes the cylindrical flow in figure 4. 
Even more important, it includes the RT-DC geometry shown in figure 7 and the fully three-dimensional cell in figure 8.
To bring out more clearly the latter aspect, we have added to figure 8 an illustration of our simulated system.
